\LARGE How coexistence mechanisms mediate temporal stability

### 2016-02-22 manuscript/manuscript.Rmd
first draft of intro paragraphs

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

@@

How coexistence mechanisms mediate temporal stability
=====================================================

### Andrew T. Tredennick, Peter B. Adler, and Frederick R. Adler

*Andrew T. Tredennick ([atredenn@gmail.com](mailto:atredenn@gmail.com)), Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

*Frederick R. Adler, Departments of Biology and Mathematics, University of Utah, Salt Lake City, UT*

@@

Introduction
------------
Species-specific responses to non-constant environmental conditions can stabilize species coexistence [@Chesson2000] and ecosystem function [@Loreau2010].
This means that fluctuation-dependent mechanisms of species coexistence are the very same mechanisms that link biodiversity and ecosystem function.
Yet, the theory that has developed over the past 20 years to explain the, generally, positive relationship between species richness and stability of ecosystem function has implicitly assumed species coexistence [@Loreau2010], or, when explicitly considered, coexistence is maintained by fluctuation-independent mechanisms [@Turnbull2012].
Despite rapid theoretical developments in the fields of species coexistence and biodiversity-ecosystem function, a gulf remains between the two lines of inquiry [@Turnbull2012].
This is especially surprising since stable coexistece, however maintained, is a preqrequisite for biodiversity to confer stability of ecosystem function in the long term.

To that end, we will analyze a general consumer-resource model under different fluctuation-dependent coexistence assmptions. Our starting point is a model of two plant consumers and one resource (e.g., soil moisture or nitrogen). We will focus on three cases of species coexistence:

\begin{enumerate}
  \item Relative nonlinearty
  \item Temporal storage effect
  \item A combination of both mechanisms
\end{enumerate}

Each scenario requires different model assumptions and structure, so we will describe each in turn. Although the structure may change slightly to incorporate different coexistence mechanisms, the strength of our approach lies in the similarities among the models since we work under a unified consumer-resource framework.

Model description and analysis
------------------------------
### A general consumer-resource model

@@

We start with a general consumer-resource model where the consumer can be in one of two-states: a dormant state $D$ and a live state $N$ (Fig. 1). Transitions between $N$ and $D$ occur at discrete intervals $T$, so our model is formulated as ``pulsed differential equations" [@Pachepsky2008; @Mailleret2009]. For clarity we refer to $T$ as years and the growing time between years, $\tau$, as seasons. Seasonal (within-year) dynamics are modeled as three differential equations:

\begin{align}
\frac{\text{d}D_{i}}{\text{d}\tau} &= -(m_{D,i}D_{i})\\
\frac{\text{d}N_{i}}{\text{d}\tau} &= N_{i}[f_{i}(R) - m_{N,i}]\\
\frac{\text{d}R_{i}}{\text{d}\tau} &= - \sum\limits_{i=1,2}f_{i}(R)N_{i}
\end{align}

where $i$ denotes species, $D$ is the dormant (long-lived) biomass state, $N$ is the living biomass (fast-growing, shorter-lived) state, and $m$s are biomass loss rates. The growth rate of living biomass is a resource-dependent Hill function, $f_{i}(R) = r_{i}R^{\alpha_{i}} / (\beta_{i}^{\alpha_{i}}+R^{\alpha_{i}})$. Resource depletion is equal to the sum of each species' consumption, $\sum_{i=1,2}f_{i}(R)N_{i}$. Note that since transitions between $N$ and $D$ are pulsed, only biomass loss occurs throughout the season for $D$.

@@

### Implementing the storage effect
To make this a ``storage-effect" model, we need to satisfy three conditions: (1) the organisms must have a mechanism for persistence under unfavorable conditions, (2) species must respond differently to environmental conditions, and (3) the effects of competition on a species must be more strongly negative in good years relative to unfavorable years. Our model meets condition 1 because we include a dormant stage with very low death rates. We satisfy condition 2 with our model whenever $g$ is not perfectly correlated between species. Lastly, our model meets condition 3 because condition 2 partitions intraspecific and interspecific competition into different years. Thus, during a high $g$ year for one species, resource uptake is also inherently high for that species, which increases intraspecific competition relative to interspecific competition. So, given adequate variability in $g$, the inferior competitor (species with lower $r$) can persist.

@@

### Implementing relative nonlinearity
When considering consumer-resource dynamics, species coexistence by relative nonlinearity requires that each species has different nonlinear responses to resource availability, and resource availability must fluctuate through time. In a constant resource environment, the species with the lowest R* will always exclude the other species. So we can compare this model to the storage effect model, we still allow the germination rate \emph{g} to vary, but both species are perfectly correlate -- that is, $\rho=1$.

Results
-------
### Storage effect model

>>>>

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in
linkcolor: black
urlcolor: black

@@

\begin{centering}

\textbf{\large{How Fluctuation-Dependent Coexistence Mechanisms Affect the Temporal Stability of Ecosystem Function}}

\end{centering}

\vspace{2.5em}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

**Andrew T. Tredennick\textsuperscript{1,}\footnote{Corresponding author; email: atredenn@gmail.com}, Peter B. Adler\textsuperscript{1}, and Frederick R. Adler\textsuperscript{2}**

\vspace{1.5em}

\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}}; \textit{\small{\textsuperscript{2}Departments of Biology and Mathematics, University of Utah, Salt Lake City, Utah}}

@@

\hrulefill

ABSTRACT: For biodiversity to increase the temporal stability of ecosystem function in the long-term, coexistence must be stable.
Species-specific respones to environmental variation through time is key to understanding fluctuation-dependent coexistence mechanisms and how asynchrony in species dynamics can arise to stabilize ecosystem function.
Despite the shared dependence on environmental fluctuations, theory on species coexistence and the relationship between species richness and ecosystem stability have developed independently.
To formally link the two bodies of theory, we use consumer-resource models where coexistence between two species utilizing a single resource is maintained by two fluctuation-dependent mechanisms: the storage effect and relative nonlinearity.
We examine how the strength of species coexistence relates to the temporal stability of aggregate ecosystem function and how the effect of environmental variability on stability is mediated by the mechanism of coexistence.
Blah, blah, blah...

*Keywords*: coexistence, storage effect, relative nonlinearity, diversity-stability hypothesis, pulsed differential equation, consumer-resource dynamics, synchrony

\hrulefill

Introduction
------------
Species-specific responses to non-constant environmental conditions can stabilize species coexistence [@Chesson2000] and ecosystem function [@Loreau2010].
This means that fluctuation-dependent mechanisms of species coexistence are the very same mechanisms that link biodiversity and ecosystem function.
Yet, the theory that has developed over the past 20 years to explain the, generally, positive relationship between species richness and stability of ecosystem function has implicitly assumed species coexistence [@Loreau2010], or, when explicitly considered, coexistence is maintained by fluctuation-independent mechanisms [@Turnbull2013].
Despite rapid theoretical developments in the fields of species coexistence and biodiversity-ecosystem function, a gulf remains between the two lines of inquiry [@Carroll2011; @Turnbull2013].
This is especially surprising since stable coexistece, however maintained, is a preqrequisite for biodiversity to confer stability of ecosystem function in the long term.

Among the dizzying array of mechanisms that can maintain diversity, @Chesson2000 formalized two broad classes: fluctuation-independent and fluctuation-dependent mechanisms.
Both classes of mechanisms rely on niche differences being greater than relative fitness differences for all species pairs in a community [@Adler2007a].
In a fluctuation-independent case, species coexistence can be maintained by resource partitioning so long as each species is limited by a different resource [@Tilman1982].
Much of the theoretical literature aimed at deciphering the mechanisms behind the diversity-stability relationship implicitly assumes fluctuation-independent coexistence.
For example, Lotka-Volterra models have been widely used and include coexistence by keeping competition coefficients less than one [e.g., @Loreau2013].
Our focus, on the other hand, is on the interaction between species coexistence and ecosystem stability in communities where coexistence is dependent on environmental fluctuations.

There are two ways that species coexist in temporally fluctuating environments.
First, there is the storage effect, where species coexistence is stable if the following three conditions are met: (i) have unique responses to environmental conditions, (ii) have some way to persist in unfavorable years, and (iii) the effects of competition must be greater in 'good' years relative to 'bad' years [@Chesson2000].
Second, there is relative nonlinearity, where species have unique, nonlinear responses to a shared resource that fluctuates through time [@Chesson2000].

Is the stability of coexistence related to the stability of ecosystem function through time?

Model and Analysis
------------------------------
### A General Consumer-Resource Model

@@

We start with a general consumer-resource model where the consumer can be in one of two-states: a dormant state $D$ and a live state $N$ (Fig. 1). Transitions between $N$ and $D$ occur at discrete intervals $T$, so our model is formulated as ``pulsed differential equations" [@Pachepsky2008; @Mailleret2009]. For clarity we refer to $T$ as years and the growing time between years as seasons with daily ($\tau$) time steps. Seasonal (within-year) dynamics are modeled as three differential equations:

\begin{align}
\frac{\text{d}D_{i}}{\text{d}\tau} &= N_{i}a_{i} - (m_{D,i}D_{i})\\
\frac{\text{d}N_{i}}{\text{d}\tau} &= N_{i}[f_{i}(R) - m_{N,i} - a_{i}]\\
\frac{\text{d}R}{\text{d}\tau} &= r_{\tau} - \sum\limits_{i=1,2}f_{i}(R)N_{i}
\end{align}

where $i$ denotes species, $D$ is the dormant (long-lived) biomass state, $N$ is the living biomass (fast-growing, shorter-lived) state, *a* is fraction of life biomass allocated to seed production, and *m* is the biomass loss rate. The growth rate of living biomass is a resource-dependent Hill function, $f_{i}(R) = b_{i}R^{\alpha_{i}} / (\beta_{i}^{\alpha_{i}}+R^{\alpha_{i}})$, where *b* is a species' intrinsic growth rate and $\alpha$ and $\beta$ define the curvature of the function. The single resource *R* is replenished with daily pulse, $r_{\tau}$, randomly drawn from a log normal distribution: $r_{\tau} = \text{LogNormal}(R_\mu, R_\sigma)$. Resource depletion is equal to the sum of each species' consumption, $\sum_{i=1,2}f_{i}(R)N_{i}$.

@@

### Implementing the Storage Effect
To make this a ``storage-effect" model, we need to satisfy three conditions: (1) the organisms must have a mechanism for persistence under unfavorable conditions, (2) species must respond differently to environmental conditions, and (3) the effects of competition on a species must be more strongly negative in good years relative to unfavorable years. Our model meets condition 1 because we include a dormant stage with very low death rates. We satisfy condition 2 with our model whenever $g$ is not perfectly correlated between species. Lastly, our model meets condition 3 because condition 2 partitions intraspecific and interspecific competition into different years. Thus, during a high $g$ year for one species, resource uptake is also inherently high for that species, which increases intraspecific competition relative to interspecific competition. So, given adequate variability in $g$, the inferior competitor (species with lower $r$) can persist.

@@

### Implementing Relative Nonlinearity
When considering consumer-resource dynamics, species coexistence by relative nonlinearity requires that each species has different nonlinear responses to resource availability, and resource availability must fluctuate through time. In a constant resource environment, the species with the lowest R* will always exclude the other species. So we can compare this model to the storage effect model, we still allow the germination rate \emph{g} to vary, but both species are perfectly correlate -- that is, $\rho=1$.

Results
-------

Discussion
----------
